Numerical simulations of the Cordilleran ice sheet through the last glacial cycle

### 2015-05-27 cordillera-cycle.tex
Add reference to Perkins and Brennand 2014.

Similarly to other glaciated regions, most glacial traces in the North American
Cordillera relate to the last few millennia of glaciation, because most of the
older evidence has been overprinted by warm-based ice retreat
during the deglaciation \citep{Kleman.1994, Kleman.etal.2010}. From a
numerical modelling perspective, phases of glacier retreat are more challenging
than phases of growth, because they involve more rapid fluctuations of the ice
margin, increased flow velocities and longitudinal stress gradients, and poorly
understood hydrological processes. The latter are typically included in the
models through simple parametrisations \citep[e.g.][]{Clason.etal.2012,
Clason.etal.2014, Bueler.Pelt.2014}, if included at all. However, next after a
matching of modelled ice margin positions with mapped end moraines
\citep{Napieralski.etal.2007}, indicative of maximum ice sheet extents
(Sect.~\ref{sec:mis2} and~\ref{sec:mis4}), geomorphologically-based
reconstructions of patterns of of ice sheet retreat during the last
deglaciation provide the second best source of evidence for the validation of
our simulations.

@@

Because a general tenant in glacial geomorphology is that the majority of
landforms (lineations and eskers) are part of the deglacial envelope
\citep[terminology from][]{Kleman.etal.2006}, having been formed
close inside the retreating margin of ice sheets \citep{Boulton.Clark.1990,
Kleman.etal.1997, Kleman.etal.2010}, we present maps of basal flow directions
immediately preceding deglaciation or at the time of cessation of sliding
inside a cold-based retreating margin (Fig.~\ref{fig:lastflow}). The
modelled deglacial flow patterns appear consistent between the GRIP and EPICA
simulations. They depict an active ice sheet retreat in the peripheral areas,
followed by stagnant ice decay in some of the interior regions. Several parts
of the modelling domain do not experience any basal sliding throughout the
deglaciation phase (Fig.~\ref{fig:lastflow}, hatched areas). This notably
includes parts of the Interior Plateau in British Columbia, major
portions of the Alaskan sector of the ice sheet, and a tortuous ribbon running
from the Northern Rocky Mountains over the Skeena and Selwyn Mountains and into
the Mackenzie Mountains.

@@

However, on the Interior Plateau of south-central British Columbia, both
simulations predict negligible basal sliding during deglaciation
(Fig.~\ref{fig:lastflow}), but an impressive set of glacial lineations
indicate a substantial eastward flow component of the Cordilleran ice sheet
\citep{Prest.etal.1968, Kleman.etal.2010}.
The Fraser lineation set could thus present a smoking gun for the
reliability of the presented model results. One explanation for the incongruent
results could therefore be that missing feedback mechanisms between ice sheet
topography and regional climate position the modelled ice divide of the LGM ice
sheet too far to the east (Sect.~\ref{sec:mis2}; Fig.~\ref{fig:mis2};
\citealp{Seguinot.etal.2014}). A more westerly-located LGM ice divide would
certainly result in a different deglacial flow pattern over the Interior
Plateau. However, a more westerly-positioned LGM ice divide would certainly be
associated with a thinner ice sheet than that modelled here. Decreased ice
thickness would not promote warm-based conditions but, on the
opposite, enlarge the region of negligible basal sliding
(Fig.~\ref{fig:lastflow}). Thus, a second explanation for the incongruent
results could be that the Interior Plateau lineation system observed in the
geomorphology predates the deglaciation, as perhaps indicated by some eskers
that appear incompatible with these glacial lineations
\citep[Fig.~9]{Margold.etal.2013a}. Finally, a third explanation could be that
local geothermal heat associated with volcanic activity on the Interior Plateau
could have triggered the basal sliding
\citep[cf. Greenland ice sheet,][]{Fahnestock.etal.2001}.

>>>>

Similarly to other glaciated regions, most glacial traces in the North American
Cordillera relate to the last few millennia of glaciation, because most of the
older evidence has been overprinted by warm-based ice retreat
during the last deglaciation \citep{Kleman.1994, Kleman.etal.2010}. From a
numerical modelling perspective, phases of glacier retreat are more challenging
than phases of growth, because they involve more rapid fluctuations of the ice
margin, increased flow velocities and longitudinal stress gradients, and poorly
understood hydrological processes. The latter are typically included in the
models through simple parametrisations \citep[e.g.][]{Clason.etal.2012,
Clason.etal.2014, Bueler.Pelt.2014}, if included at all. However, next after a
matching of modelled ice margin positions with mapped end moraines
\citep{Napieralski.etal.2007}, indicative of maximum ice sheet extents
(Sect.~\ref{sec:mis2} and~\ref{sec:mis4}), geomorphologically-based
reconstructions of patterns of of ice sheet retreat during the last
deglaciation provide the second best source of evidence for the validation of
our simulations.

@@

Because a general tenant in glacial geomorphology is that the majority of
landforms (lineations and eskers) are part of the deglacial envelope
\citep[terminology from][]{Kleman.etal.2006}, having been formed
close inside the retreating margin of ice sheets \citep{Boulton.Clark.1990,
Kleman.etal.1997, Kleman.etal.2010}, we present maps of basal flow directions
immediately preceding deglaciation or at the time of cessation of sliding
inside a cold-based retreating margin (Fig.~\ref{fig:lastflow}). The
modelled deglacial flow patterns are mostly consistent between the GRIP and EPICA
simulations. They depict an active ice sheet retreat in the peripheral areas,
followed by stagnant ice decay in some of the interior regions. Several parts
of the modelling domain do not experience any basal sliding throughout the
deglaciation phase (Fig.~\ref{fig:lastflow}, hatched areas). This notably
includes parts of the Interior Plateau in British Columbia, major
portions of the Alaskan sector of the ice sheet, and a tortuous ribbon running
from the Northern Rocky Mountains over the Skeena and Selwyn Mountains and into
the Mackenzie Mountains.

@@

On the Interior Plateau of south-central British Columbia, both simulations
produce a retreat of the ice margin towards the north-east
(Fig.~\ref{fig:deglac}), a pattern which is validated by the geomorphological
and stratigraphical record for ancient pro-glacial lakes dammed by the
retreating ice sheet \citep{Perkins.Brennand.2014}. However, the two
simulations differ in the mode of retreat. The GRIP simulation yields an active
retreat with basal sliding towards the ice margin to the south, whereas the
EPICA simulation produce negligible basal sliding during deglaciation
(Fig.~\ref{fig:lastflow}). Yet, the Interior Plateau also hosts an impressive
set of glacial lineations which indicate a substantial eastward flow component
of the Cordilleran ice sheet \citep{Prest.etal.1968, Kleman.etal.2010}.
The Fraser lineation set could thus present a smoking gun for the
reliability of the presented model results. One explanation for the incongruent
results could therefore be that missing feedback mechanisms between ice sheet
topography and regional climate position the modelled ice divide of the LGM ice
sheet too far to the east (Sect.~\ref{sec:mis2}; Fig.~\ref{fig:mis2};
\citealp{Seguinot.etal.2014}). A more westerly-located LGM ice divide would
certainly result in a different deglacial flow pattern over the Interior
Plateau. However, a more westerly-positioned LGM ice divide would certainly be
associated with a thinner ice sheet than that modelled here. Decreased ice
thickness would not promote warm-based conditions but, on the
opposite, enlarge the region of negligible basal sliding
(Fig.~\ref{fig:lastflow}). Thus, a second explanation for the incongruent
results could be that the Interior Plateau lineation system observed in the
geomorphology predates the deglaciation, as perhaps indicated by some eskers
that appear incompatible with these glacial lineations
\citep[Fig.~9]{Margold.etal.2013a}. Finally, a third explanation could be that
local geothermal heat associated with volcanic activity on the Interior Plateau
could have triggered the basal sliding
\citep[cf. Greenland ice sheet,][]{Fahnestock.etal.2001}.
